Simulating High-Dimensional Multivariate Data

### 2020-12-01 index.Rmd
3rd draft. run spell checking and edit

Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
Further, efficient simulation empowers statistical computing strategies, such as the parameteric bootstrap [@Rizzo2007] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
Such Monte Carlo (MC) techniques become difficult for high-dimensional data with the current existing algorithms and tools.
This is particularly true when simulating massive *multivariate*, *non-normal* distributions, arising naturally in many fields of study.

As others have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19].
For continous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula [@Nelsen2007] approaches are well-studied with flexible, robust software available [@Yan2007; @Chen2001].
Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu]. 
For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible dependencies (e.g., admitting only positive correlations @Park1996). 
While more recent approaches [@MB13; @Xia17; @BF17] have largely rememdied this issue for low-dimesional problems, the existing tools are not designed to scale to high dimensions.

@@

With all this mind, we present a scalable, flexible multivariate simulation algorithm. 
The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software implemented our NORTA-inspirsed algorithm.
The algorithm design leverages useful properties of nonparameteric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.

>>>>

Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry.
As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness --- often in comparison to existing methods.
Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Rizzo2007] to simulate from a hypothesized null model, providing inference in analytically challenging settings.
Such Monte Carlo (MC) techniques become difficult for high-dimensional data with the current existing algorithms and tools.
This is particularly true when simulating massive *multivariate*, *non-normal* distributions, arising naturally in many fields of study.

As others have noted, it can be vexing to simulate dependent, non-normal/discrete data ---  even for low dimensional settings [@MB13; @XZ19].
For continuous non--normal multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula [@Nelsen2007] approaches are well-studied with flexible, robust software available [@Yan2007; @Chen2001].
Yet these approaches do not scale in a timely fashion to high-dimensional problems [@Li2019gpu]. 
For discrete data, early simulation strategies had major flaws --- such as failing to obtain the full range of possible dependencies (e.g., admitting only positive correlations @Park1996). 
While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for low-dimensional problems, the existing tools are not designed to scale to high dimensions.

@@

With all this mind, we present a scaleable, flexible multivariate simulation algorithm. 
The crux of the method lies in the construction of a Gaussian copula, in the spirit of the NORTA procedure.
Further, we introduce the `bigsimr` R package that provides parallelized, high-performance software implemented our NORTA-inspired algorithm.
The algorithm design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed form relationships between dependency measures for the multivariate normal (MVN) distribution.
